Think again?\\ Optimal mental simulation tracks problem difficulty

### 2015-01-25 manuscript.tex
More work on introduction

%% The question we're interested in: how many simulations do people take?
Imagine you are playing a video game---for example, Angry Birds or Desert Golf---in which you need to predict where an object will end up. Recent research in the domain of intuitive physics and mental simulation has suggested that people make predictions like these by running noisy physical simulations \cite{Smith2012,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}. However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. For example, if people are running mental simulations, \emph{how many} simulations do they run?

\TODO{something discussing that there are scenarios in which it's useful to take more simulations than others?}

%% What's already been done
Previous research by \citeA{Vul:2014ba} has asked, in the abstract, how many samples an agent should take before making a decision. Surprisingly, they found that to optimize success in the long run, it is not useful for sample-based agents to take more than a few samples. For example, suppose that there is a reward for correct answers and no penalty for incorrect answers, and that the agent can take ten samples in the time it takes to actually execute the action that reflects the decision. Then, the optimal number of samples that the agent should base their decision on is only one! While this is obviously suboptimal for each individual decision, over many decisions, this strategy trades off between the amount of time one can actually spend taking actions versus thinking.

%% What's missing
So far, the theoretical results from \citeA{Vul:2014ba} have largely not been applied to the domain of mental simulation, and lead to several questions regarding how people use their mental simulations. First, do people run multiple mental simulations? If so, do they run a fixed number of simulations, or vary the number of simulations based on the difficult? A supplementary analysis by \citeA{Battaglia2013} suggests that people do run multiple simulation. In their analysis, they attempted to determine the number of samples taken by people in several tasks concerning towers of building blocks. They did find a consistent story with \citeA{Vul:2014ba}: participants in their experiments seemed to use between three and seven samples per judgment. However, their analysis only concerned the variability of responses and did not take into account response time.

%% What we're going to do and how we're going to do it
In this paper, we investigate how many mental simulations people run when making decisions in a two alternative force choice task, analyzing both their response times and the variability of responses. \TODO{plan of the paper}

>>>>

%% The question we're interested in: how many simulations do people take?
Imagine you are playing a video game---for example, Angry Birds or Desert Golf---in which you need to predict where an object will end up. How do you know where to aim, and how long do you spend thinking about each shot before you take it? Recent research in the domain of intuitive physics and mental simulation has revealed that people make predictions like these by running noisy physical simulations \cite{Smith2012,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}. However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. If people are running mental simulations, \emph{how many} simulations do they run?

%% What's already been done
According to previous research by \citeA{Vul:2014ba}, a sample-based agent should only take a small number of samples before making a judgment---perhaps even just one. While taking a single sample is clearly suboptimal for each individual judgment, over the long run, taking a small number of samples maximizes expected utility across a large number of judgments. Does this story also hold true for mental simulations? There is some evidence that points to ``yes'': \citeA{Battaglia2013} analyzed the variability of people's responses in tasks concerning towers of building blocks, and found that participants seemed to use between three and seven samples per judgment.

A second question concerns the actual strategy used to take samples. Do people always take exactly the same number of samples for each decision, or do they vary the number of samples based on problem difficulty? While \citeA{Vul:2014ba} looked at a variety of strategies, they concluded that regardless of strategy, the optimal solution was still roughly the same: the agent should take very few samples. However, that does not make the choice of strategy irrelevant. If the strategy tells the agent to vary the number of samples, then we should see variability in the agent's response times. Indeed, such variability can be observed in other types of decision-making tasks, and has been well-modeled by the \textit{sequential probability ratio test} \cite{wald1947sequential}, or SPRT \cite<e.g.>{Gold:2007fo,Ratcliff:2008ux,Bitzer:2014ea}. In SPRT, the agent takes samples until it reaches a threshold number of samples in support of one hypothesis over another hypothesis.

%% What we're going to do and how we're going to do it
In this paper, we investigate the number of simulations that people run when reasoning about a task that requires mental simulation of physics. Specifically, we ask the question: do people vary the number of mental simulations that they run depending on the difficulty of the problem? And, if so, do they do so in a manner consistent with SPRT? To answer these questions, we ran a behavioral experiment in which people watched a ball bouncing around in a box, and had to judge whether it would first go through a hole in the wall, or bounce off that wall. We varied the difficulty of each trial by changing the size of the holes and the margin by which the ball either went through the hole or missed the hole. We hypothesized that people should be faster to respond on ``easy'' trials (e.g., where the ball goes directly through the center of a large hole, or hits the wall very far away from the hole), and slower to respond on ``hard'' trials (e.g., where the ball just barely goes through the hole, or just barely misses the hole). In brief, find that people's responses are well-predicted by a model of simulation, and that their response times are well-predicted by combining that model with a SPRT strategy.

%% What we're going to do and how we're going to do it
The plan of the paper is as follows. First, we formalize our hypothesis for how people respond to the question, ``will the ball go through the hole?'', specifying both a simulation-based model as well as the SPRT strategy. Second, we describe two behavioral experiments, one in which we asked people to respond to the question of ``will the ball go through the hole?'', and a related experiment that we used to determine the parameters for the simulation-based model. Third, we describe the results of the experiments, and compare people's judgments and response times to those from the model. Finally, we discuss the implications of our results on the broader, underlying question: how do people use mental simulation?
